Review
======
 
```{r cache=TRUE}
#Integer (L)
x = 1L
y = 1
#integer
class(x) 
#numeric
class(y)
```

Logical
```{r}
TRUE
FALSE
```


Vectors
same class
```{r cache=TRUE}
heights = c(188.2, 182, 193)
names = c('jeff', 'roger', 'andrew', 'brian')

```

lists
```{r cache=TRUE}
mylist = list(heights=heights, firstnames=names)
mylist
mylist$heights
mylist$firstnames

```

matrices
* vectors with multiple dimensions
* left to right by row, 2 rows
```{r cache=TRUE}
mymatrix = matrix(c(1, 2, 3, 4), byrow=T, nrow=2)
mymatrix
```


data frames - multiple vectors possibly of different class, same length
```{r cache=TRUE}
df = data.frame(heights=c(heights, 199), firstnames=names)
df
```

factors - qualitative variables that can be included in models
```{r cache=TRUE}
smoker = c('yes', 'no', 'yes', 'yes')
smokerFactor = as.factor(smoker)
smokerFactor
```

missing values - NA
```{r cache=TRUE}
vector1 = c(199.2, 291, 192.5, NA)
is.na(vector1)
```

subsetting
```{r cache=TRUE}
heights = c(heights, 192.3)
df = data.frame(heights=heights, firstNames=names)
```

just first value of heights
```{r}
heights[1]
```

indices
```{r}
heights[c(1, 2, 4)]
```

first row, and columns 1 to 2
```{r}
df[1, 1:2]
df$firstNames
df[df$firstNames=='jeff']
df[df$heights < 190,]
```

Simulation
==========
distributions
* rfoo functions generate data
```{r}
heights = rnorm(10, mean=188, sd=3)
```

function(n, size, prob)
* n - number of scenarios
* size - number of coins "flipped"
* prob - each coins is fair, number of heads is added up
```{r}
rbinom(10, size=10, prob=0.5)
```

densities - probability of X values being in a some ranges - function(x, mean=0, sd=1, log=FALSE) log units
```{r}
x = seq(from=5, to=5, length=10)
```

calc normal density of those points
```{r}
normalDensity = dnorm(x, mean=0, sd=1)
round(normalDensity, 2)
```

normal density - function(x, size, prob, log=FALSE)
```{r}
x = seq(0, 10, by=1)
binomialDensity = dbinom(x, size=10, prob=0.5)
round(binomialDensity, 2)
```


sampling - function(x, size, replace=FALSE, prob=NULL)
```{r}
heights = rnorm(10, mean=188, sd=3)
```

sample with replacement may sample the same value, draw one random value from vector of heights, then replaces that value in vector of heights and then draws a new random sample from among the 10 values
```{r}
sample(heights, size=10, replace=T)
```

value removed from heights vector, and so forth, in this case the we end up with the same vector because sizes are the same
```{r}
sample(heights, size=10, replace=F)
```

sampling in accordance with probabilities - 0.4 prob that first value sampled, etc
```{r}
probs = c(0.4, 0.3, 0.2, 0.1, 0, 0, 0, 0, 0, 0)
sample(heights, size=10, replace=T, prob=probs)
```

setting seed - allows generating same random values
```{r}
set.seed(12345)
rnorm(5, mean=0, sd=1)
```

